---
title: "Using the New York Times Books API in R"
author: "Guibril Ramde"
format: html
editor: visual
---

## Approach

For this assignment, I selected the \*\*New York Times Books API\*\*, specifically the endpoint for the current \*\*hardcover fiction best seller list\*\*. The purpose of this analysis is to retrieve data from the API, convert the JSON response into a tidy data frame in R, and examine the current best-selling books on the list.\
\
To access the API securely, I created a New York Times Developer Network account and generated an API key. Instead of hard coding the key directly in the script, I stored it in my \`.Renviron\` file and accessed it in R using \`Sys.getenv("NYT_API_KEY")\`. This approach protects the key from being exposed in a public GitHub repository.\
\
Next, I used the \`httr\` package to make a GET request to the Books API endpoint. After receiving the response, I used \`jsonlite\` to parse the JSON content. Since the full API response contains nested fields and metadata, I extracted the \`books\` portion from \`data\$results\`, which contains the actual book records for the selected list.\
\
I then transformed the nested JSON data into a tidy tibble and selected the variables most relevant for analysis, such as rank, title, author, publisher, and weeks on list. Any unnecessary nested fields were excluded in order to produce a cleaner and more readable data frame.\
\
Finally, I used the cleaned data to perform a small analysis of the current hardcover fiction best sellers and included a simple table or visualization to help answer an interesting question based on the API results.

## Coding Section

```{r}
library(httr)
library(tidyverse)
library(jsonlite)
library(tibble)
```

```{r}
api_key <- Sys.getenv("NYT_API_KEY")
api_key

url <- paste0("https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json?api-key=", api_key)

response <- GET(url)
response
data <- fromJSON(content(response, "text"))

data
```
